---
title: "eQTL mapping with QuasarExperiment"
author: "Vincent Carey"
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document:
    toc: true
    toc_float: true
vignette: >
  %\VignetteIndexEntry{eQTL mapping with QuasarExperiment}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>"
)
```

# Overview

`QuasarExperiment` provides a
[SummarizedExperiment](https://bioconductor.org/packages/SummarizedExperiment)-style
container for expression quantitative trait loci (eQTL) mapping with the
[quasar](https://github.com/jeffreypullin/quasar) C++ tool.  quasar implements
a wide range of models — linear models (LM), negative binomial GLMs (NB-GLM),
and linear mixed models (LMM) — and is generally faster than comparable tools.

The key design principle is that large genotype data (PLINK `.bed` files) are
represented **lazily** via
[BEDMatrix](https://CRAN.R-project.org/package=BEDMatrix): the binary file is
never loaded into R memory.  Phenotype values, covariates, and variant metadata
are held in the familiar SE slots.

# Installation

```{r install, eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("vjcitn/QuasarExperiment")
```

On macOS ARM64 (Apple Silicon) a pre-built quasar binary is bundled in the
package and used automatically.  On other platforms, build quasar from source
(see `?runQuasar`) or set `options(QuasarExperiment.binary = "/path/to/quasar")`.

# Data

This vignette uses a small example dataset bundled in `inst/extdata`: 20 genes
and 69,638 variants on chromosome 22, subsetted to 100 individuals from the
OneK1K B cell naive cluster.  Phenotype values are pseudobulk mean counts.

```{r load}
library(QuasarExperiment)

exdir <- system.file("extdata", package = "QuasarExperiment")
dir(exdir)
```

# Constructing a QuasarExperiment

`QuasarExperiment()` reads the phenotype BED file, covariate TSV, and PLINK
file set.  The optional `grmFile` argument adds a genetic relatedness matrix
for use with the LMM model.

```{r construct}
qe <- QuasarExperiment(
    plinkPrefix = file.path(exdir, "chr22-n100"),
    phenoFile   = file.path(exdir, "mean-pheno-n100.bed"),
    covFile     = file.path(exdir, "cov-n100.tsv"),
    grmFile     = file.path(exdir, "grm-n100.tsv")
)
qe
```

The object reports its dimensions (features × samples), the names of loaded
assays, the number of variants in the lazy genotype matrix, and whether a GRM
is present.

# Accessing slots

`QuasarExperiment` inherits all `SummarizedExperiment` accessors.

## Phenotype assay

```{r assay}
# 20 genes x 100 samples
dim(assay(qe, "pheno"))
assay(qe, "pheno")[1:3, 1:4]
```

## Gene ranges

```{r rowranges}
rowRanges(qe)
```

## Sample metadata and covariates

Covariates from the TSV file are stored as columns of `colData()`.

```{r coldata}
head(colData(qe)[, c("int", "sex", "age", "expr_pc1", "geno_pc1")])
```

## Lazy genotype matrix

`geno()` returns a `BEDMatrix` — a proxy object backed by the PLINK `.bed`
file.  Dimensions are reported immediately; data are only read when subscripted.

```{r geno}
geno(qe)                     # 100 samples x 69638 variants
geno(qe)[1:3, 1:5]           # materialise a small slice
```

## Variant ranges

Variant coordinates and allele information are stored as a `GRanges`.

```{r variants}
variantRanges(qe)
```

## Genetic relatedness matrix

```{r grm}
dim(grm(qe))
grm(qe)[1:4, 1:4]
```

# Running quasar

`runQuasar()` writes temporary input files, invokes the quasar binary, and
returns the results as `GRanges`.  The `model` and `mode` arguments map
directly to quasar's `--model` and `--mode` flags.

On macOS ARM64 the bundled binary is used automatically; the currently bundled
version can be checked with `bundledQuasarVersion()`.

```{r version}
bundledQuasarVersion()
```

## Linear model (cis)

```{r lm, eval=nzchar(Sys.which("quasar")) || .Platform$OS.type != "windows"}
res_lm <- runQuasar(qe, model = "lm", mode = "cis")
```

## Inspecting results

`variantHits()` returns one range per feature–variant pair tested.
`regionHits()` returns one range per feature (the genome-wide minimum p-value).

```{r results, eval=exists("res_lm")}
variantHits(res_lm)
regionHits(res_lm)
```

Sort by p-value to find the strongest associations:

```{r tophits, eval=exists("res_lm")}
vh <- variantHits(res_lm)
vh[order(vh$pvalue)][1:10]
```

## Negative binomial GLM

For count-based phenotypes (sum pseudobulk) the NB-GLM with the Cox-Reid
adjusted profile likelihood (`useApl = TRUE`) is recommended.

```{r nb, eval=FALSE}
qe_sum <- QuasarExperiment(
    plinkPrefix = file.path(exdir, "chr22-n100"),
    phenoFile   = file.path(exdir, "sum-pheno-n100.bed"),
    covFile     = file.path(exdir, "cov-n100.tsv"),
    assayName   = "counts"
)
res_nb <- runQuasar(qe_sum, model = "nb_glm", mode = "cis", useApl = TRUE)
```

## Linear mixed model

The LMM accounts for population structure via the GRM.

```{r lmm, eval=FALSE}
res_lmm <- runQuasar(qe, model = "lmm", mode = "cis")
```

# Session info

```{r sessioninfo}
sessionInfo()
```
